Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-12-20 arxiv/main.tex
update for thea's comments 2

\begin{abstract}
A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on a global biosphere disturbance. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. Crucially, this augmentation must be accompanied by a causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people. We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
\end{abstract}

@@

\section*{Introduction}
Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}. 
A large body of research reveals how these models fit the mortality seen due to the Coronavirus epidemic in 2020 \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.

@@

This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, a conclusion made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?

@@

While Schumann Resonance disturbances can be caused by natural geomagnetic events, others have looked at possible causative links through man-made influences on the ionosphere and magnetosphere through the use of electromagnetic infrastructure \cite{siingh2005review}. It is thus conceivable to have an anthropogenic disturbance of the ionosphere through the use of e.g. satellite communication technologies. Not only might this effect the
global electric circuit \cite{rycroft2008overview} and the Schumann Resonance, but the usage of satellite communication technologies might also alter the magnetosphere itself leaving a signal similar to Power Line Harmonic Radiation \cite{bullough2017power,kositsky2001influence}.

@@

Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics. Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model. This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations. In doing this we shall show from a physical argument that the Gompertz model is incompatible with disease spreading through social interactions.

@@

 In general, the SIR models and their extensions which do not consider network interactions explicitly consider rather averaged macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond}. In such models, all entities are connected and communicating through long-range interactions as shown by Mombach et al. (2002) \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIR model family predicts, but rather a constant negative slope in the double-log-domain, which we will now show.

\subsection*{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease models. Under a double-log-transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve \cite{Levitt2020},
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This extension requires an addition to our original set of equations \ref{eq:SIR}, with an equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
\begin{equation}
\frac{dD}{dt} = c D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in Eq. \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$. Both the SIR and SEIR model exhibit similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \cite{vattay2020forecasting}.

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \cite{carletti2020covid}. The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \cite{efron2012bayesian}. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

>>>>

\begin{abstract}
A number of studies have shown that cumulative mortality followed a Gompertz curve in the initial covid pandemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on a global biosphere disturbance. 
Second, we show that for the Gompertz curve to emerge, the disturbance has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model family is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. Crucially, this augmentation must be accompanied by a causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people. We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
\end{abstract}

@@

\section*{Introduction}
Traditional communicable disease spreading theory assumes a pathogen which infects the population through a network of transmission. 
Following this line of reasoning it can be theoretically shown that in the early stages of an epidemic, growth follows a logistic-like curve, where the very beginning exhibits exponential growth, and for which analytic and semi-analytic solutions have recently emerged \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately}. 
A large body of research reveals how these models fit the mortality seen due to the Covid pandemic in 2020, a pandemic due to an influenza-like disease \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}.

@@

This phenomenon is recorded by others \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the predicted logistic curves, an observation made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. What is causing such a discrepancy between these observations and the current theory of communicable diseases \cite{castro2020turning}?

@@

While Schumann Resonance disturbances can be caused by natural geomagnetic events, others have looked at possible causative links through man-made influences on the ionosphere and magnetosphere through the use of electromagnetic infrastructure \cite{siingh2005review}. It is thus conceivable to have an anthropogenic disturbance of the ionosphere through the use of e.g. satellite communication technologies. Not only might this effect the
global electric circuit \cite{rycroft2008overview} and the Schumann Resonance, but the usage of satellite communication technologies might also alter the magnetosphere itself leaving a signal similar to Power Line Harmonic Radiation \cite{bullough2017power}.

@@

Whatever the cause, global biosphere disturbance theory represents an alternative paradigm in disease spreading dynamics. Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model. This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations. In doing this we shall show that the Gompertz model is incompatible with disease spreading through physical proximity interactions. Further, we show that the Gompertz curve emerge when individuals are connected through a common stressor affecting them simultaneously.

@@

 In general, the SIR models and their extensions which do not consider network interactions explicitly consider rather averaged macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond}. In such models, all entities are implicitly assumed connected and communicating through long-range interactions as shown by Mombach et al. (2002) \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIR model family predicts, but rather a constant negative slope in the double-log-domain, which we will now show.

\subsection*{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richards model, one can still almost obtain Gompertzian growth from the communicable disease SIR-like models. Under a double-log-transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve \cite{Levitt2020},
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR model by including a group of deceased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This extension requires an addition to our original set of equations \ref{eq:SIR}, with an equation for the deceased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
\begin{equation}
\frac{dD}{dt} = c D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in Eq. \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$. Both the SIR and SEIR model exhibit similar properties and their discrepancy with initial observations related to the Covid pandemic has been noted by others \cite{vattay2020forecasting}.

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Covid pandemic, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \cite{carletti2020covid}. The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \cite{efron2012bayesian}. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}
